Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-15 notebooks/expected_and_actual_counts_ratio.Rmd
Group by amino acid and add counts of identical mutations

# Identify the two clades that will be included in comparisons.
clades <- c("20A","22B")

@@

# Analyze nucleotide mutations or amino acid mutations ('nt' or 'aa')?
option <- "nt"
merge_column <- paste(option, "mutation", sep="_")
```

## Filter and Prepare the SARS-CoV-2 Mutation Data

@@

```{r aa_mutation}
if(option=='aa'){
  aa_group <- filtered_SARS_mutation
  filtered_SARS_mutation <- as.data.frame(aa_group %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
}else{
  nt_group <- filtered_SARS_mutation
  filtered_SARS_mutation <- nt_group[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","actual_count","expected_count")]
}

@@

@@

```{r head_data}
head(filtered_SARS_mutation, n=5)
```

```{r summary_data}
summary(filtered_SARS_mutation)
```

Take a closer look at how many nucleotide sites and the proportion of mutations in the data set.

```{r describe_data}
# How many nucleotide sites?
length(unique(filtered_SARS_mutation$nt_site))

# What percentage of the data set is each type of nucleotide mutation? (e.g., A -> T)
nt_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
(table(nt_type)/length(nt_type))*100

# What percentage of the data set is each type of amino acid mutation? (e.g., K -> M)
aa_type <- paste(str_sub(filtered_SARS_mutation$clade_founder_aa,1,1), ".", str_sub(filtered_SARS_mutation$mutant_aa,1,1), sep="")
aa_type <- paste(filtered_SARS_mutation$clade_founder_aa, ".", filtered_SARS_mutation$mutant_aa, sep="")
(table(aa_type)/length(aa_type))*100
```

```{r stacked_nucleotide, echo=FALSE}
stacked_nt <- filtered_SARS_mutation
stacked_nt$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")

```{r stacked_amino_acid, echo=FALSE}
stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(79)

stacked_aa <- filtered_SARS_mutation
#stacked_aa$clade_founder <- str_sub(filtered_SARS_mutation$clade_founder_aa,1,1)
#stacked_aa$mutant <- str_sub(filtered_SARS_mutation$mutant_aa,1,1)

ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant_aa))+
  geom_bar(position="stack", stat="identity")+
  facet_wrap(~clade_founder_aa, ncol=7)+
  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
  scale_x_discrete(guide=guide_axis(n.dodge=5))+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

## Restructure Data for Analyses

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
#p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))

# Save and load the variable because calculating p-values may take a long time.
#save(p_values, file=paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
load(paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
```

@@

```{r volcano_corrected, echo=FALSE}
# After dropping P = 1, plot the corrected p-values.
color <- colorRampPalette(brewer.pal(8,"Set1"))(comparison_p_values[,paste("clade_founder", option, sep="_")] %>% unique() %>% length())

volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=get(paste("clade_founder", option, sep="_"))))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()
```

>>>>

# Identify the two clades that will be included in comparisons.
clades <- c("22A","22B")

@@

# Analyze nucleotide mutations or amino acid mutations ('nt' or 'aa')?
option <- "aa"
merge_column <- paste(option, "mutation", sep="_")
```

## Filter the SARS-CoV-2 Mutation Data

@@

Create nucleotide mutation and amino acid mutation data sets. Actual and expected counts for rows with identical amino acid mutations are added together. Both data sets only include columns relevant for that mutation to make understanding restructuring the data and analyses easier.

```{r group_mutation}
# Group by amino acid mutation and add counts for identical amino acid mutations.
# NOTE: Grouping with and without 'gene' creates a different number of rows because some equivalent aa mutations belong to different genes. There are 2,160 aa mutations that have duplicates.
aa_group <- as.data.frame(filtered_SARS_mutation %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous, gene) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))

# Reduce columns for nucleotide mutation analyses to make workflow easier to follow.
nt_group <- filtered_SARS_mutation[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","gene","actual_count","expected_count")]

@@

```{r head_data}
head(get(paste(option, "group", sep="_")), n=5)
```

```{r summary_data}
summary(get(paste(option, "group", sep="_")))
```

Take a closer look at how many nucleotide sites, amino acid sites and the proportion of mutations in the data set.

```{r describe_data}
# How many nucleotide sites?
length(unique(nt_group$nt_site))

# How many amino acid sites?
length(unique(aa_group$codon_site))

# What percentage of the data set is each type of nucleotide mutation? (e.g., A -> T)
nt_type <- paste(str_sub(nt_group$nt_mutation,1,1), ".", str_sub(nt_group$nt_mutation,-1,-1), sep="")
(table(nt_type)/length(nt_type))*100

# What percentage of the data set is each type of amino acid mutation? (e.g., K -> M)
aa_type <- paste(aa_group$clade_founder_aa, ".", aa_group$mutant_aa, sep="")
(table(aa_type)/length(aa_type))*100
```

```{r stacked_nucleotide, echo=FALSE}
stacked_nt <- nt_group
stacked_nt$mutation_type <- paste(str_sub(nt_group$nt_mutation,1,1), ".", str_sub(nt_group$nt_mutation,-1,-1), sep="")

```{r stacked_amino_acid, echo=FALSE}
stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(21)

stacked_aa <- aa_group
stacked_aa$clade_founder <- str_sub(aa_group$clade_founder_aa,1,1)
stacked_aa$mutant <- str_sub(aa_group$mutant_aa,1,1)

ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant))+
  geom_bar(position="stack", stat="identity")+
  facet_wrap(~clade_founder, ncol=7)+
  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
  scale_x_discrete(guide=guide_axis(n.dodge=5))+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

## Prepare and Restructure Data for Analyses

```{r rename_variable}
# Rename variable for analyses to make code more readable.
filtered_SARS_mutation <- get(paste(option, "group", sep="_"))
```

@@

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))

# Save and load the variable because calculating p-values may take a long time.
#save(p_values, file=paste("../results/expected_and_actual_counts_ratio/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
load(paste("../results/expected_and_actual_counts_ratio/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
```

@@

```{r volcano_corrected, echo=FALSE}
# After dropping P = 1, plot the corrected p-values.
volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
if("gene" %notin% colnames(volcano)){
  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep=";"))
}
color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))

p <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()

p + theme(legend.position="none")
```

```{r volcano_corrected_legend}
p + theme(legend.position="bottom")
```
